---
title: "Reproducible Research: Peer Assessment 1"
author: "MnkyKng"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
1. Loading and first investigation of the dataset:  
```{r echo=TRUE}
dataset <- "activity.csv"

dataset_folder <- if (!file.exists("activity.csv")){
        unzip("activity.zip")
}

dataset <- read.csv("activity.csv")
dim(dataset)

head(dataset)
mean(is.na(dataset))
```
Our dataset has 17568 rows and 3 columns. It also contains 4.37% NA (missing) values.  

2. Removal of the NA values from the dataset:  
```{r echo=TRUE}
library(dplyr)
filtered_dataset <- dataset %>% filter(!is.na(steps))
str(filtered_dataset)
```


## What is mean total number of steps taken per day?
1. Calculation of total number of steps taken per day:    
```{r echo=TRUE}
filtered_dataset$date <- as.Date(filtered_dataset$date)
str(filtered_dataset)

steps_per_day <- filtered_dataset %>% group_by(date) %>% summarise(sum(steps)) %>% rename(total_steps='sum(steps)')
head(steps_per_day)
```

2. Histogram of total number of steps taken per day:   
```{r echo=TRUE}
library(ggplot2)

g <- ggplot(steps_per_day, aes(total_steps))
g+geom_histogram()+ggtitle("Distribution of total steps taken per day (NA values removed)")+xlab("Total steps taken per day")+ylab("Frequency")
```

3. Mean and Median of total number of steps taken per day:   
```{r echo=TRUE}
mean(steps_per_day$total_steps)
median(steps_per_day$total_steps)
```
We have a mean of 10766 steps and a median of 10765 steps.


## What is the average daily activity pattern?
1. Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):  
```{r echo=TRUE}
average <- tapply(dataset$steps, dataset$interval, mean, na.rm=TRUE)
summary(is.na(average))

df_iavg <- data.frame(interval=as.integer(names(average)), avg=average)
summary(df_iavg)
        
with(df_iavg, plot(interval, avg, type = "l", main = "Average number of steps taken, averaged across all days", xlab = "Interval", ylab = "Average number of steps"))
```

2. Calculation of the 5-minute interval containing the maximum number of steps, on average across all the days in the dataset:  
```{r echo=TRUE}
max_steps <- max(df_iavg$avg)
df_iavg[df_iavg$avg == max_steps, ]
```
The interval at 835 contains the maximum number of 206 steps.


## Imputing missing values
1. The total number of missing values in the dataset (TRUE = missing values):  
```{r echo=TRUE}
summary(is.na(dataset))
```
Our dataset contains 2304 total missing values (NA's).

2. Filling in of all the missing values in the dataset,   
3. and creation of a new dataset:   
```{r echo=TRUE}
library(tidyr)

new_dataset <- dataset %>% replace_na(list(steps=0))
str(new_dataset)

summary(is.na(new_dataset))
```
Replaced all NA values in columns "steps" with 0. Compared with the previous summary, we don't have any is.na = TRUE anymore. That means, because our NA values are now 0, they have been added to steps = FALSE.  

4. Histogram of total number of steps taken per day based on the new_dataset:
```{r echo=TRUE}
new_dataset$date <- as.Date(new_dataset$date)
new_dataset$steps <- as.integer(new_dataset$steps)
str(new_dataset)

steps_per_day_new <- new_dataset %>% group_by(date) %>% summarise(sum(steps)) %>% rename(total_steps_new ='sum(steps)')
head(steps_per_day_new)

a <- ggplot(steps_per_day_new, aes(total_steps_new))
a+geom_histogram()+ggtitle("NEW distribution of total steps taken per day (NA values replaced with 0)")+xlab("Total steps taken per day")+ylab("Frequency")

mean(steps_per_day_new$total_steps_new)
median(steps_per_day_new$total_steps_new)
```
Due to the number of new values (NA's have been replaced with 0, not removed), our new mean dropped to 9354 steps as well as our median shifted to 10395 steps.


## Are there differences in activity patterns between weekdays and weekends?
1. Creation of a factor variable with two levels indicating a "weekday" or a "weekend":  
```{r echo=TRUE}
new_dataset$datetype <- sapply(new_dataset$date, function(x){
        if(weekdays(x) == "Saturday" | weekdays(x)=="Sunday"){
                datetype <- "weekend"
        } else {
                datetype <- "weekday"
        }
})
head(new_dataset)
summary(new_dataset$datetype=="weekend")
```
We have a total amount of 12960 weekdays and 4608 weekends.

2. Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday or weekend (y-axis):  
```{r echo=TRUE}
aggregated_data <- aggregate(steps~interval + datetype, new_dataset, mean)
g <- ggplot(aggregated_data, aes(x=interval, y=steps, color=datetype))
g+geom_line()+facet_wrap(~datetype, nrow = 2, ncol = 1)+ggtitle("Average number of steps taken per day by type of date")+xlab("Interval")+ylab("Average number of steps")
```